### 2014-07-10 writeups/context-dependence.tex
A few more changes to make things clearer IMHO.

Alternatively, we could take into account context in the initial sequence only,
by including the states of an additional $\ell$ sites on both the left and right.
The probability that an initial sequence $x$ of length $m+2\ell$ after time $t$
is found to match a smaller, middle sequence $y$ of length $m$ is found by marginalizing:
\begin{align}
  p_{\ell,m}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^\ell} p_{2\ell+m}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{2\ell+m}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $2\ell+m$,
then after time $t$ we will see at the middle $m$ positions the pattern $y$,
ignoring what is present in the leftmost and rightmost $\ell$ positions.
We expect $p_{\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
  p_{\ell,m}(x,y) \approx \P\{ X_{i+\ell}^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
\end{align}
We provide a precise statement of the error in this approximation later.

@@

If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
(using Stein's method)
Then we'll have the correct likelihood function for (this subset of) data.
If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
and any overlapping patterns must require additional changes,
i.e.\ if $(x_1,y_1)$ matches at $i_1$,
then for $(x_2,y_2)$ to match at $i_2$,
there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.

Here is one way to obtain such a set.
Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
For example, in the example above with $m=3$ and $\ell=2$,
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
$x$ and $y$ differ only at (relative) position $i_1=2$.
Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
As another example,
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  A & C & T & C & A & C & G \\
 $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
has $i_1 = \bar i = 1$.
No two such patterns can be shifted versions of each other
without adding more changed sites,
since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.

>>>>

Alternatively, we could take into account context in the initial sequence only,
by including the states of an additional $\ell$ sites on both the left and right.
The probability that an initial sequence $x$ of length $m+2\ell$ after time $t$
is found to match a smaller, middle sequence $y$ of length $m$ is found by marginalizing:
\begin{align}
  p_{\ell,m}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^\ell} p_{2\ell+m}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{2\ell+m}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $2\ell+m$,
then after time $t$ we will see the pattern $y$ at the middle $m$ positions,
ignoring what is present in the leftmost and rightmost $\ell$ positions.
We expect $p_{\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
  p_{\ell,m}(x,y) \approx \P\{ X_{i+\ell}^{(m)}(t) = y \mid X_i^{(2\ell+m)}(0) = x \}.
\end{align}
We provide a precise statement of the error in this approximation later.

@@

If we use only a subset of patterns $\{(x_k,y_k)\}$ chosen so that $\E[M_1(x,y) M_{k+1}(x,y)]$ is small (or zero),
the entire collection $\{N(x_k,y_k)\}$ will be close to independent Poisson.
(using Stein's method)
Then we'll have the correct likelihood function for (this subset of) data.

If changes are rare, then one way to get a set of patterns with this property is to only choose ones such that going from $x$ to $y$ entails a change;
and any overlapping patterns must require additional changes,
i.e.\ if $(x_1,y_1)$ matches at $i_1$,
then for $(x_2,y_2)$ to match at $i_2$,
there must be additional changes outside $(i_1,\ldots,i_1+\ell-1)$.

Here is one way to obtain such a set.
Take a pair of patterns $(x,y)$, that overlap at $m$ positions,
and suppose that they differ at $1 \le i_1, \ldots, i_d \le m$.
Let $\bar i = (1/d) \sum_{j=1}^d i_j$ be the mean position of the changes,
and take only patterns with $(m-1)/2 < \bar i \le (m+1)/2$.
For example, in the example above with $m=3$ and $\ell=2$,
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  C  & A & C & T & C & A & C \\
 $y$ &  $\centerdot$  & $\centerdot$  & C & G & C & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
$x$ and $y$ differ only at (relative) position $i_1=2$.

Another way to achieve the same goal of having all overlapping patterns require additional changes is to require $i_1 = \bar i = 1$.
No two such patterns can be shifted versions of each other
without adding more changed sites,
since shifting by $k$ without adding changed sites just adds $k$ to $\bar i$.
For example, we might take
\begin{center}
\begin{tabular}{c|ccccccc}
 $x$ &  A & C & T & C & A & C & G \\
 $y$ &  $\centerdot$  & $\centerdot$  & G & C & A & $\centerdot$  & $\centerdot$
\end{tabular}
\end{center}
